---
output: 
  html_document: 
    self_contained: yes
    keep_md: yes
---

# APAstats

For automated and basic inferential testing, lets you perform:

-   Student's or Welch's *t*-test
-   Mann–Whitney *U* test
-   Yuen's test for trimmed means
-   Fisher's or Welch's one-way ANOVA
-   Kruskal–Wallis one-way ANOVA
-   Heteroscedastic one-way ANOVA for trimmed means
-   Wilcoxon signed-rank test
-   Yuen's test on trimmed means for dependent samples
-   One-way repeated measures ANOVA
-   Friedman rank sum test
-   Heteroscedastic one-way repeated measures ANOVA for trimmed means

## Installation

For installation of developmental version run in your R console:

``` r
install.packages("devtools")
devtools::install_github('matcasti/APAstats')
```

and then restart your R session.

## Automated testing

By default, `report` function, checks automatically the assumptions of the test based on the parameters of the data entered

```{r echo=TRUE}
library(APAstats) # Load the APAstats package

plant <- datasets::PlantGrowth # Let's use the PlantGrowth package
                               # (installed by default in R)

result <- report(
  data = plant,
  variable = 'weight', # dependent variable
  by = 'group', # independent variable
  pairwise.comp = TRUE, # for pairwise comparisons in case of n groups > 2
  markdown = TRUE # output in Markdown format (for inline text)? otherwise plain text.
)

result
```

## Inline results in APA style

The core function `report` by default return a list of length two in Markdown format (as seen before), for inline results. An example using same data as before:

The analysis of the effects of the treatment shows an statistically significant difference between the groups, `result$report`, evaluated through `result$method`.

translates into this:

The analysis of the effects of the treatment shows an statistically significant difference between the groups, `r result$report`, evaluated through `r result$method`.


## Citation

To cite package 'APAstats' in publications run the following code in your `R` console:

```{r}
citation('APAstats')
```

## Dependencies

The package APAstats is standing on the shoulders of giants:

```{r fig.height=10, fig.width=10}
library(deepdep)
plot_dependencies('APAstats', local = TRUE, depth = 3)
```

## Contact

For issues or collaborations on the package you can send me an email at:

-   matcasti\@umag.cl
